An Introduction to Rocker: Docker Containers for R

### 2017-07-10 manuscript.Rmd
first (incomplete) pass with typo edits
minimal style / wording / content changes

---
title: "An Introduction to Rocker: Docker Containers for R"
author:
  - name: Carl Boettiger
    affiliation: UC Berkeley 
    address:
    - ESPM Department, University of California,
    - 130 Mulford Hall Berkeley, CA 94720-3114, USA
    email:  cboettig@berkeley.edu
  - name: Dirk Eddelbuettel
    affiliation: Debian and R Projects
    address:
    - line 1
    - line 2
    email:  edd@debian.org
abstract: >
  We describe the Rocker project, which provides a widely used suite of Docker images
  with customized R environments for particular tasks. We discuss how this suite is
  organized and how these tools can increase portability, scaling, reproducibility,
  and convenience of R users and developers.
preamble: >
   \usepackage{longtable}
   \usepackage{svg}
output: rticles::rjournal_article
---

@@

Docker is a popular open source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment requiring everything an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight (sharing the operating system kernel, starting instantly,
using a layered filesystem to minimize disk footprint and download time),
built on open standards that run on all major platforms (Linux, Mac,
Windows) and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.

@@

The Rocker Project consists of a suite of images built automatically by
and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.  Source
Dockerfiles, supporting scripts and documentation are hosted on GitHub
under the organization `rocker-org`, <https://github.com/rocker-org>.
The issues tracker and pull requests are used to provide community input,
disucssions and contributions to these images. The rocker project wiki
provides a place to synthesize community-contributed documentation,
use-cases, and other knowledge about using the Rocker images.

@@

The Rocker Project aims to provide a small core if Docker images that
serve as convenient 'base' images on which other users can build custom
R environments by writing their own Dockerfiles, while also providing a
'batteries included' approach of images that can be used out of the box.
The challenges of balancing diverse needs driven by very different use
cases (discussed in more detail below) against the overarching goals
of creating images that are not sufficiently light-weight, easy to use
and easy to maintain is a difficult art.  The implementation in both
individual Rocker images and image stacks can never perfect that balance
for everyone, but today reflects the considerable community input and
testing over past few years.

All Rocker images are based on the Debian Linux Distribution.
The Debian platform provides a small base image, the well known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
Unfortunately, the relatively long period between stable Debian releases
(roughly two years recently) means that software in the `debian:stable`
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.  The Rocker Project
can be largely divided into two stacks which address different needs,
defined by which Debian distribution they are based on.  The first stack
is based on `debian:testing`, the second, more recently introduced stack
is based only on `debian` stable releases.  These stacks have different
aims and thus provide different images, as shown in Tables 1 & 2.

@@

The `debian:testing` stack aims to makes the most efficient use of
upstream builds -- the precompiled `.deb` binaries provided by the
Debian repositories.  It is both quicker and easier to install software
from binaries, since the package manager (`apt`) manages the necessary
dependencies and it bypasses the time-consuming process of compiling
from source.  Basing this stack on `debian:testing` means that much more
recent versions of commonly used libraries and compilers are available
as binaries than would be found in a Debian stable release.  In order
to provide access to the most recent avialble binaries, this stack uses
apt-pinning [@apt-pinning] to allow the `apt` package manager to also
install binaries from `debian:unstable`, which represents the most recent,
bleeding edge of packages built for Debian when necessary.  For instance,
the `r-base` image provided by Rocker installs the most recent version of
R as a binary from Debian `unstable`.  Similarly, recent versions of many
popular R packages can also be installed through the package manager,
e.g. `apt-get install r-cran-XML`.  This can be particularly helpful
for packages with external system dependencies (such as `libxml2-dev`
in this example) which cannot be installed from the R console.

As the names `testing` and `unstable` imply, this approach is not
without a downside.  The particular version of any given package can
change as packages move from `unstable` into `testing` and newer versions
are sent to `unstable` during the normal course of Debian development.
This can occassionally break an previously working installation command
in a Dockerfile until the maintainer redirects the package manager to
install a the package from the `unstable` sources that could previously
be installed from `testing`, or vice versa (using the `-t` option in
`apt`, see Box 1 on apt-pinning.)

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility [@apt-pinning].
4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.

>>>>

---
title: "An Introduction to Rocker: Docker Containers for R"
author:
  - name: Carl Boettiger
    affiliation: UC Berkeley 
    address:
    - ESPM Department, University of California,
    - 130 Mulford Hall Berkeley, CA 94720-3114, USA
    email:  cboettig@berkeley.edu
  - name: Dirk Eddelbuettel
    affiliation: Debian and R Projects; Ketchum Trading
    address: 
    - Chicago, IL, USA
    email:  edd@debian.org
abstract: >
  We describe the Rocker project, which provides a widely-used suite of Docker images
  with customized R environments for particular tasks. We discuss how this suite is
  organized and how these tools can increase portability, scaling, reproducibility,
  and convenience of R users and developers.
preamble: >
   \usepackage{longtable}
   \usepackage{svg}
output: rticles::rjournal_article
---

@@

Docker is a popular open source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment requiring all operating-system components an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight as they share the operating system kernel, starting instantly,
using a layered filesystem which minimizes disk footprint and download time,
and are built on open standards that run on all major platforms (Linux, Mac,
Windows), and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.

@@

The Rocker Project consists of a suite of images built automatically by
and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.  Source
Dockerfiles, supporting scripts and documentation are hosted on GitHub
under the organization `rocker-org`, <https://github.com/rocker-org>.
The issues tracker and pull requests are used for community input,
disucssions and contributions to these images. The rocker project wiki
provides a place to synthesize community-contributed documentation,
use-cases, and other knowledge about using the Rocker images.

@@

The Rocker Project aims to provide a small core of Docker images that
serve as convenient 'base' images on which other users can build custom
R environments by writing their own Dockerfiles, while also providing a
'batteries included' approach of images that can be used out of the box.
The challenges of balancing diverse needs driven by very different use
cases (discussed in more detail below) against the overarching goals
of creating images that are still sufficiently light-weight, easy to use
and easy to maintain is a difficult art.  The implementation in both
individual Rocker images and image stacks can never perfect that balance
for everyone, but today reflects the considerable community input and
testing over past few years.

All Rocker images are based on the Debian Linux Distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
Unfortunately, the relatively long period between stable Debian releases
(roughly two years recently) means that software in the `debian:stable`
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.  The Rocker Project
can be largely divided into two stacks which address different needs,
defined by which Debian distribution they are based on.  The first stack
is based on `debian:testing`, the second, more recently introduced stack
is based only on `debian` stable releases.  These stacks have different
aims and thus provide different images, as shown in Tables 1 & 2.

@@

The `debian:testing` stack aims to makes the most efficient use of
upstream builds -- the precompiled `.deb` binaries provided by the
Debian repositories.  It is both quicker and easier to install software
from binaries, since the package manager (`apt`) manages the necessary
(binary) dependencies and bypasses the time-consuming process of compiling
from source.  Basing this stack on `debian:testing` means that much more
recent versions of commonly-used libraries and compilers are available
as binaries than would be found in a Debian stable release.  In order
to provide access to the most recent avialble binaries, this stack uses
apt-pinning [@apt-pinning] to allow the `apt` package manager to also
install binaries from `debian:unstable`, which represents the most recent,
bleeding edge of packages built for Debian when necessary.  For instance,
the `r-base` image provided by Rocker installs the most recent version of
R as a binary from Debian `unstable`.  Similarly, recent versions of many
popular R packages can also be installed through the package manager,
e.g. `apt-get install r-cran-xml`.  This can be particularly helpful
for packages with external system dependencies (such as `libxml2-dev`
in this example) which cannot be installed from the R console as they 
are system dependencies rather than R packages installed from within R.

As the names `testing` and `unstable` imply, this approach is not
without challenges.  The particular version of any given package can
change as packages move from `unstable` into `testing`. New versions
are sent to `unstable` during the normal course of Debian development.
This can occassionally break an previously working installation command
in a Dockerfile until the maintainer redirects the package manager to
install a the package from the `unstable` sources that could previously
be installed from `testing`, or vice versa (using the `-t` option in
`apt`, see Box 1 on apt-pinning.)

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility [@apt-pinning].
4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.
<!-- DE: Are you sure about that point 4 ? -->
